---
title: "Statistical Modelling II"
author: Mauricio Garcia Tec
date: "`r paste('Version:', Sys.Date())`"
site: bookdown::bookdown_site
output: 
  bookdown::html_book
  bookdown::gitbook
github-repo: mauriciogtec/StatsModellingII
documentclass: book
---

# Welcome

Placeholder



<!--chapter:end:index.Rmd-->


# List 1: Preliminaries

Placeholder


## Exchangeability and de Finetti's theorem
### De Finneti's theorem
## Exponential family of distributions
### Cumulants and moments of exponential families
### Conjugate priors
## Multivariate Normal Distribution
### Manipulation of multivariate normals
## Frequentist estimation and uncertainty quantification

<!--chapter:end:01-list1.Rmd-->


# List 2: Bayesian Inference in Gaussian Models

Placeholder


## Bayesian inference in a simple Gaussian model

<!--chapter:end:02-list2.Rmd-->

---
title: "Gibbs Part III"
author: "Mauricio Garcia Tec"
date: "February 21, 2018"
output: html_document
---


```{r}
library(readr)
library(knitr) 
library(# to visualize data nicely
```


```{r}
data <- read_csv("https://raw.githubusercontent.com/sinead/sds383d/master/data/pima.csv")
kable(data)
```

```{r}
hyperparams = list(
  beta0 = numeric(ncol(data)), 
  kappa0 = 1.0, 
  a0 = 2.0, 
  b0 = 2.0
)
```

```{r}
gibbs_sampler <- 
```


<!--chapter:end:gibbs_3.Rmd-->

